Enabling Ambient Backscatter \\Using a Low-Cost Software Defined Radio

### 2017-05-15 abstract.tex
Smaller changes. References to github look better now.
Several small corrections

Backscatter communication enables ultra-low
power wireless transmissions, and is attractive
for networking devices which operate 
on harvested energy. Ambient backscatter
takes the concept further, by leveraging
ambient wireless signals like
television signals, as both the source of power
and carrier signal. Existing state-of-the-art ambient backscatter
systems demonstrate ability to achieve tag-to-tag communication 
under conditions when the
strength of TV signals  sufficiently high (-30 dBm).  In this paper, we present our preliminary work
which demonstrates the possibility to backscatter and communicate even when the ambient television signals
are significantly weaker. The key to achieving this is to leverage a low-cost
software defined radio receiver (RTLSDR) to receive backscattered transmissions. Our results 
demonstrate that the  television signals are strong 
enough in most parts of a mid-sized swedish city for our system
to operate, and communicate, a significant improvement over the
state-of-the-art which are
restricted to operate only when the tags are close to the TV towers.

>>>>

Backscatter communication enables ultra-low
power wireless transmissions, and is attractive
for networking devices which operate 
on harvested energy. Ambient backscatter
takes the concept further, by leveraging
ambient wireless signals like
television signals, as both the source of power
and carrier signal. Existing state-of-the-art ambient backscatter
systems demonstrate the ability to achieve tag-to-tag communication 
under conditions when the
strength of TV signals is sufficiently high (-30 dBm).  In this paper, we present our preliminary work
which demonstrates the possibility to backscatter and communicate even when the ambient television signals
are significantly weaker. The key to achieving this is to leverage a low-cost
software defined radio receiver (RTL-SDR) to receive backscattered transmissions. Our results 
demonstrate that the  television signals are strong 
enough in most parts of a mid-sized swedish city for our system
to operate, and communicate, a significant improvement over the
state-of-the-art which are
restricted to operate only when the tags are close to the TV towers.

### 2017-05-15 backscatter.tex
Several small corrections

Consider an unmodulated carrier wave impinging on a backscatter tag's
antenna. The signal observed by the receiver is:
\begin{equation}
				S_r(t) = S_{rc}(t) + \sigma B(t)S_{bc}(t)
				\label{eq:signals}
\end{equation}
where $S_{rc}(t)$ is the signal coming directly from the carrier
generator and $S_{bc}(t)$ is the signal from the carrier generator that
reaches the backscatter tag. $B(t)$ is either zero or one and represents
the instantaneous state of the backscatter tag: absorbing or reflecting,
respectively.

@@

Recent work on generating backscatter transmissions has avoided
self-interference through \textit{frequency-shifted}
backscatter~\cite{kellogg_passive_2016,wang_fm_2017,varshney2016lorea}.
The tags modulate their antenna in such a way that their transmissions
occur at a certain frequency offset from the carrier signal thus
allowing the receiver to avoid interference from the carrier by tuning
to the offset frequency where the tag is transmitting.

>>>>

Consider an unmodulated carrier wave impinging on a backscatter tag's
antenna. In the presense of an umodulated carrier. The signal observed
by the receiver is: \begin{equation}
				S_r(t) = S_{rc}(t) + \sigma B(t)S_{bc}(t)
				\label{eq:signals}
\end{equation}
where $S_{rc}(t)$ is the signal coming directly from the carrier
generator and $S_{bc}(t)$ is the signal from the carrier generator that
reaches the backscatter tag. $\sigma$ is the tag's radar cross section
and $B(t)$ is either zero or one and represents the instantaneous state
of the backscatter tag: absorbing or reflecting, respectively.

@@

Recent work on generating backscatter transmissions has avoided
self-interference through \textit{frequency-shifted}
backscatter~\cite{kellogg_passive_2016,wang_fm_2017,varshney2016lorea,
iyer2016inter}.
The tags modulate their antenna in such a way that their transmissions
occur at a certain frequency offset from the carrier signal thus
allowing the receiver to avoid interference from the carrier by tuning
to the offset frequency where the tag is transmitting.

### 2017-05-15 eval.tex
Smaller changes. References to github look better now.
Several small corrections

In this section, we present the results of evaluation of our system. As experimental 
setup, we use RTLSDR together with signal processing algorithm implemented
in Octave.

@@

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T is specified up to 10 MHz bandwidth.
And the local TV provider advertises to be transmitting DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\biggl( \Re\{ s_{\text{band,n}}(t) \} \biggr) \Biggr|
\end{equation}     where \ensuremath{N} is the total number of intervals
accumulated over the entire frequency sweep. The FFT is only carried out over
the real values, which are received from the \textit{RTL-SDR}.
\ensuremath{s_{\text{band,n}}(t)} is the signal in the time domain of one
interval (one subband). There are as many samples taken in one band as needed
for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to
normalize everything and calculate the power. Before doing that, a simple
smoothing algorithm (FIR) is applied on \ensuremath{|H|}.  
\begin{equation}
	|H_{\text{tv,avg}}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain point
in space. The result of combining measured signal strength values with the
haversine function to calculate the distance to the point where the maximum
signal has been captured can be observed in Figure \ref{fig:haversine}.
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\vspace{-6mm}
	\caption{\emph{Spatial variation of TV signals in Uppsala city}. 
	The left image demonstrates the signal strength of TV signal fading	
	as opposed to observed signal strength near the TV tower. 
	%474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
%594 MHz belongs to a tower in Uppsala, Rickomberga. 
	The right image visualises the signal strength as heatmap for the signal
	present in the \SI{626}{\mega\hertz} band. Distances have been
calculated with the Haversine formula. }
 	\vspace{-6mm}

@@

\subsection{Communication Performance}
We tested the communication with different data rates from 1 bit/s up to
1 kbit/s. As can be seen in Figure \ref{fig:transmission} the signal
strength from the backscatter transmitter is rather low. This leads to a
significant amount of quantization noise to appear. This situation can
be improved by increasing the analog gain of the receiver. The
transmission shown in Figure \ref{fig:transmission} was carried out with
the maximum gain (50 dB) available to our receiver.

>>>>

In this section, we present the results of evaluation of our system. As experimental 
setup, we use RTLSDR together with a signal processing algorithm implemented
in Octave.

@@

During the scanning process the gain is set to 40.2 dB. To get the
signal strength we simply take the average over a 10 MHz band around the
center frequency of the desired TV signal. This approach can be
justified by the fact, that DVB-T 2 is specified up to 10 MHz bandwidth.
And the local TV provider advertises to be transmitting DVB-T 2.
The signal \ensuremath{|H|} is calculated as follows.  
\begin{equation}
	|H| = \Biggl| \sum_{n=0}^N FFT\biggl( \Re\{ s_{\text{band,n}}(t) \} \biggr) \Biggr|
\end{equation}     where \ensuremath{N} is the total number of intervals
accumulated over the entire frequency sweep. The FFT is only carried out over
the real values, which are received from the \textit{RTL-SDR}.
\ensuremath{s_{\text{band,n}}(t)} is the signal in the time domain of one
interval (one subband). There are as many samples taken in one band as needed
for an FFT of size 2048. To finally get \ensuremath{|H_{tv,avg}|} one has to
normalize everything and calculate the power. Before doing that, a simple
smoothing algorithm (FIR) is applied on \ensuremath{|H|}.  
\begin{equation}
	|H_{\text{tv,avg}}| = 20 \cdot log (|H|) - 20 \cdot log(|H_0|)
\end{equation}   
where \ensuremath{|H_0|} is a normalization value measured at a certain point
in space. The result of combining measured signal strength values with the
haversine function to calculate the distance to the point where the maximum
signal has been captured can be observed in Figure \ref{fig:haversine}.
\begin{figure}[h]
	\centering
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/haversine}
	\end{minipage}
	\hfill
	\begin{minipage}{0.49\columnwidth}
	\includegraphics[width=\columnwidth]{./fig/heatmap_626mhz}
	\end{minipage}
	\vspace{-6mm}
	\caption{\emph{Spatial variation of TV signals in Uppsala city}. 
	The left image demonstrates the signal strength of TV signal fading	
	as opposed to observed signal strength near the TV tower. 
	%474 MHz and 626 MHz belong to a tower in Uppsala, Vedyxa and
%594 MHz belongs to a tower in Uppsala, Rickomberga. 
	The right image visualises the signal strength as heatmap for the signal
	present in the \SI{626}{\mega\hertz} band. Distances have been
calculated with the Haversine formula. }
 	\vspace{-6mm}

@@

\subsection{Communication Performance}
We tested the communication with different data rates from 1 bit/s up to
1 kbit/s. As can be seen in Figure \ref{fig:transmission} the signal
strength from the backscatter transmitter is rather low. This leads to the appearance of a
significant amount of quantization noise. This situation can
be improved by increasing the analog gain of the receiver. The
transmission shown in Figure \ref{fig:transmission} was carried out with
the maximum gain (50 dB) available to our receiver.

### 2017-05-15 introduction.tex
Several small corrections

Backscatter communication enables wireless transmissions
at energy consumption which is several orders of magnitude
lower than traditional radios. Backscatter achieves ultra-low
power wireless transmissions by reflecting or absorbing ambient
wireless signals, which as an operation consumes only 
\SI{}{\micro\watt}s of power consumption~\cite{liu_ambient_2013}. 
As a consequence backscatter
communication is emerging as the mechanism of choice to network
devices operating on harvested energy.  
Over the past few years there has been significant progress
to make backscatter a viable mechanism to network Internet of
Things~(IoT) devices. Traditional backscatter systems, like RFID readers, required a
dedicated device to generate the necessary carrier signal
reflected by the tags. On the other hand, state-of-the art systems 
does away the need for a dedicated device to generate carrier signals. 
Recent backscatter systems leverage already deployed infrastructure of
devices to generate carrier signal~\cite{varshney2016lorea,iyer2016inter}, or ambient WiFi~\cite{hitchhike,kellogg2015wi} or TV
signals~\cite{liu_ambient_2013,parks_turbocharging_2014}.

Recent backscatter systems demonstrate ability to leverage
existing signals like TV signals as a both source of carrier and energy. For
example, Liu et al. present a proof-of-concept system that reflects
ambient TV signals and enables tag-to-tag communication up to almost a
meter. Parks et al. further improve the communication range to several
meters by using analog coding~\cite{parks_turbocharging_2014}. While
these systems can enable many applications, these systems are severely restricted
to operate in the vicinity of television towers where ambient signal
are sufficiently strong~(approx
\SI{-30}{dBm}). This is primarily due to poor sensitivity levels of receivers
employed on these devices, which together with weak backscatter reflections 
severely limits the operating range from the tower.
On the other hand, TV signals are known to vary greatly in
strength~\cite{wang_fm_2017} both over space, and in time,  which further aggravates the
problem of limited range of ambient backscatter systems.

@@

In this paper we explore the following questions: Can we leverage an
SDR-based receiver to receive ambient backscatter transmissions ?, and, Does the  
relatively high sensitivity of SDR receivers improve
range and coverage of ambient backscatter systems? A positive answer to the above question would provide a
flexible and low-cost experimentation platform to the wider research community to explore ambient backscatter
systems.

@@

				\item We design the first system which leverages a  low-cost 
					  SDR to receive ambient-backscatter transmissions. 
				\item  Using the system designed, we demonstrate ambient backscatter using TV signals
					   to be feasible in wide parts of a city. The range represents a significant improvement
					   over the state-of-the-art.

@@

>>>>

Backscatter communication enables wireless transmissions
at an energy consumption several orders of magnitude
lower than traditional radios. Backscatter achieves ultra-low
power wireless transmissions by reflecting or absorbing ambient
wireless signals, which consumes only 
\SI{}{\micro\watt}s of power~\cite{liu_ambient_2013}. 
As a consequence backscatter
communication is emerging as the mechanism of choice to network
devices operating on harvested energy.  
Over the past few years there has been significant progress
to make backscatter a viable mechanism to network Internet of
Things~(IoT) devices. Traditional backscatter systems, like RFID readers, required a
dedicated device to generate the necessary carrier signal
reflected by the tags. On the other hand, state-of-the art systems does away
with the need for a dedicated device to generate carrier signals.  Recent
backscatter systems leverage the already deployed infrastructure such as
WiFi~\cite{hitchhike,kellogg2015wi} or TV
towers~\cite{liu_ambient_2013,parks_turbocharging_2014} to generate the carrier
signal. %~\cite{varshney2016lorea,iyer2016inter}, or ambient WiFi.

Recent backscatter systems demonstrate the ability to leverage
existing signals, such as TV, as the source of both carrier and energy. For
example, Liu et al. present a proof-of-concept system that reflects
ambient TV signals and enables tag-to-tag communication up to almost a
meter. Parks et al. further improve the communication range to several
meters by using analog coding~\cite{parks_turbocharging_2014}. While
these systems can enable many applications, they are severely restricted
to operate in the vicinity of television towers where ambient signal
are sufficiently strong~(approx
\SI{-30}{dBm}). This is primarily due to the poor sensitivity of the receivers
employed on these devices, which together with weak backscatter reflections 
severely limits the operating range from the tower.
Furthermore, TV signals are known to vary greatly in
strength~\cite{wang_fm_2017} both over space, and in time,  which further aggravates the
problem of the limited range of ambient backscatter systems.

@@

In this paper we explore the following questions: Can we leverage an
SDR-based receiver to receive ambient backscatter transmissions?, and, Does the  
relatively high sensitivity of SDR receivers improve the
range and coverage of ambient backscatter systems? A positive answer to the above questions would provide a
flexible and low-cost experimentation platform to the wider research community to explore ambient backscatter
systems.

@@

				\item We design the first system that leverages a  low-cost 
					  SDR to receive ambient-backscatter transmissions. 
				\item  We demonstrate ambient backscatter using TV signals
					   to be feasible in wide parts of a city. The range represents a significant improvement
					   over the state-of-the-art.

@@

The paper proceeds as follows. First, we discuss relevant background information of backscatter communication,
SDR and reception process. Next, we present the design of the transmitter and the receiver employed in this
paper. We then present initial results which evaluates our system. Finally, we present concluding remarks.

### 2017-05-15 paper.tex
Fixed wrong figure reference a reviewer pointed to
Several small corrections

\begin{figure}[h]
				\centering
				\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
				\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q
				demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner
				with integrated LNA \textit{Rafael Micro R820T/2} (right IC).} 
				\label{fig:receiver_arch}
\end{figure}
The \textit{Realtek RTL2832U} is a terrestrial digital video broadcast
(DVB-T) demodulator IC, which is the main part of a wide range of
USB-based DVB-T receiver dongles. One prominent feature of this chip is
that it allows to retrieve the raw I/Q samples via USB. This is
originally intended for the chip to work as a simple DAB/FM software
receiver. Ham radio enthusiasts have combined their efforts to create a
software driver (the \textit{librtlsdr}, cf. \cite{steve-m_librtlsdr})
that allows DVB-T dongles based on the \textit{RTL2832} to be converted
into low-cost wideband SDR receivers. The cost of traditional SDRs has
been in the range of a few hundred to thousands of USD. Albeit much less
powerful than a typical SDR, a \textit{RTL2832U}-based DVB-T receiver
can be bought with antenna for less than 10 USD.

An RTL-SDR also contains a tuner that allows the user to select the
received frequency.  In our case the tuner is a \textit{Rafael Micro
R820T/2} which offers a tuning range from 42 to 1002
MHz~\cite{rafael_r820t_2011}. Figure \ref{fig:receiver_arch}, shows a
photograph of the RTL-SDR used in our work.

@@

Backscatter operates by reflecting or absorbing impinging
radio signals on the antenna. To purposefully toggle the antenna
to achieve these states, we  terminate the antenna to a 
RF switch. The RF switch enables us to alternate the impedance connected to
the antenna between matched (50 \ensuremath{\Omega}) and open state.
We control the RF switch through an I/O pin on a low-powered MCU, Texas
instruments MSP430. As discussed earlier, to reduce the effects of
self-interference, we frequency shift the backscatter transmission
away from the TV signal. Hence, we operate the RF switch using a 
\SI{2}{\mega\hertz} intermediate frequency signal. Therefore the result 
is a frequency shift of the television signal by 2 MHz, when
transmitting a bit 1, and no shift for a 0.  Hence,  we 
leverage amplitude modulation at receiver in our design.
binary amplitude shift keying is the implemented modulation technique.

@@

The resulting values are sent through the bus to the decoder. After the decoder
received the samples it decides when a frame is sent or when the channel is
idle. The decoder also includes a function to correlate the received bitstream
with an expected pattern. Hence it is able to determine the bit error ratio
(BER). Furthermore the code we have written so far subsumes different
simulators for e.g. playing back recorded data. Another handy utility, which
has been written in Octave, is a tool (an oscilloscope) to look at the data,
which is written into files by the demodulator. An example of a recorded data
stream with 25 kS/s can be seen in Figure \ref{fig:transmission}.

@@

To come to a conclusion one can say, that we were able to show, that
spectrum scanning over a huge band with sweeping is possible using the
\textit{RTL-SDR}.  This allowed us to perform a survey of the TV signal
strength in the city. And finally we have shown for the first time, that
backscatter communication is possible with the \textit{RTL-SDR}.

>>>>

\begin{figure}[h]
				\centering
				\includegraphics[width=0.5\columnwidth]{./fig/rtlsdr.jpg}
				\caption{\textit{RTL-SDR} hardware with the DVB-T I/Q
				demodulator \textit{Raeltek RTL2832U} (left IC) and the tuner
				with integrated LNA \textit{Rafael Micro R820T/2} (right IC).} 
				\label{fig:rtl_hw}
\end{figure}
The \textit{Realtek RTL2832U} is a terrestrial digital video broadcast
(DVB-T) demodulator IC, which is the main component in a wide range of
USB-based DVB-T receiver dongles. One prominent feature of this chip is
that it allows to retrieve the raw I/Q samples via USB. This is
originally intended for the chip to work as a simple DAB/FM software
receiver. Ham radio enthusiasts have combined their efforts to create a
software driver (the \textit{librtlsdr}~\cite{steve-m_librtlsdr})
that allows DVB-T dongles based on the \textit{RTL2832} to be converted
into low-cost wideband SDR receivers. The cost of traditional SDRs is
in the range of a few hundred to thousands of USD. Albeit much less
powerful than a typical SDR, a \textit{RTL2832U}-based DVB-T receiver
can be bought with antenna for less than 10 USD.

An RTL-SDR also contains a tuner that allows the user to select the
received frequency.  In our case the tuner is a \textit{Rafael Micro
R820T/2} which offers a tuning range from 42 to 1002
MHz~\cite{rafael_r820t_2011}. Figure \ref{fig:rtl_hw}, shows a
photograph of the RTL-SDR used in our work.

@@

Backscatter operates by reflecting or absorbing impinging
radio signals on the antenna. To purposefully toggle the antenna
to achieve these states, we  terminate the antenna to a 
RF switch. The RF switch enables us to alternate the impedance connected to
the antenna between matched (50 \ensuremath{\Omega}) and open state.
We control the RF switch through an I/O pin on a low-powered MCU, Texas
instruments MSP430. As discussed earlier, to reduce the effects of
self-interference, we frequency-shift the backscatter transmission
away from the TV signal. Hence, we operate the RF switch using a 
\SI{2}{\mega\hertz} intermediate frequency signal. Therefore the result 
is a frequency shift of the television signal by 2 MHz, when
transmitting a bit 1, and no shift for a 0.  Hence,  we 
leverage amplitude modulation at receiver in our design.
binary amplitude shift keying is the implemented modulation technique.

@@

The resulting values are sent through the bus to the decoder. After the decoder
received the samples it decides when a frame is sent or when the channel is
idle. The decoder also includes a function to correlate the received bitstream
with an expected pattern. Hence it is able to determine the bit error ratio
(BER). Furthermore we have implemented a series of aditional tools that include
different simulators for e.g. playing back recorded data and a tool (an
oscilloscope) to plot the data, which is written into files by the demodulator.
An example of a recorded data stream with 25 kS/s can be seen in Figure
\ref{fig:transmission}.

@@

%To come to a conclusion one can say, that 
We have shown, that
spectrum scanning over a wide band with sweeping is possible using the
\textit{RTL-SDR}.  This allowed us to perform a survey of the TV signal
strength in the city. We found that the signal strength around most of the city
is adequate for backscatter communication using our receiver. And finally we
have shown for the first time, that backscatter communication is possible with
the \textit{RTL-SDR}.
